---
title: "faissR: nearest neighbours, graphs, and clustering"
author:
- name: "Stefano Cacciatore"
package: faissR
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{faissR: nearest neighbours, graphs, and clustering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 6,
    fig.height = 4
)
```

# Overview

`faissR` provides native nearest-neighbour search, k-nearest-neighbour
prediction, and k-means utilities for R
workflows that use large row-wise matrices. The package is designed for
high-throughput biological data analysis, including single-cell, flow
cytometry, imaging, and mass spectrometry applications.

The package requires the FAISS C++ library for CPU indexes. CUDA, FAISS GPU
indexes and NVIDIA cuVS are optional. CPU-only
installations do not need NVIDIA libraries, and explicit CUDA requests fail
clearly when CUDA support is not available.

The main public functions are:

| Function | Purpose |
|---|---|
| `nn()` | nearest-neighbour search on host matrices |
| `nn_gpu()` | GPU-resident exact-family KNN result for downstream CUDA code |
| `candidate_knn()` | refine a supplied candidate neighbour set |
| `knn()` and `predict()` | supervised kNN fitting and prediction |
| `fast_kmeans()` | CPU or optional CUDA k-means |
| `backend_info()` and `nn_capabilities()` | inspect compiled backends |

# Example data

The examples use a small synthetic matrix with three groups. Rows are
observations and columns are features, matching the input convention used by
`faissR`.

```{r data}
library(faissR)

set.seed(100)
n_per_group <- 30L
p <- 8L

centers <- rbind(
    rep(-2, p),
    rep(0, p),
    rep(2, p)
)

x <- do.call(rbind, lapply(seq_len(nrow(centers)), function(i) {
    sweep(
        matrix(rnorm(n_per_group * p, sd = 0.7), ncol = p),
        2,
        centers[i, ],
        "+"
    )
}))

groups <- factor(rep(LETTERS[1:3], each = n_per_group))
dim(x)
table(groups)
```

# Backends

`backend_info()` reports which compiled routes are available in the current R
session. On a CPU-only machine, CUDA-related entries are expected to be
unavailable.

```{r backends}
backend_info()[, c("backend", "available", "knn_available")]
```

`nn_capabilities()` gives method, backend, and metric support. The table can be
large, so the example below shows only CPU Euclidean rows.

```{r capabilities}
caps <- nn_capabilities()
subset(
    caps,
    backend == "cpu" & metric == "euclidean",
    select = c("method", "backend", "metric", "supported", "exact")
)
```

# Nearest-neighbour search

The main entry point is `nn()`. Use `exclude_self = TRUE` for self-KNN when
each row is searched against the same matrix and should not return itself.

```{r nn-exact}
nearest <- nn(
    x,
    k = 5,
    backend = "cpu",
    method = "exact",
    metric = "euclidean",
    exclude_self = TRUE,
    n_threads = 2
)

nearest
dim(nearest$indices)
head(nearest$indices)
head(round(nearest$distances, 3))
```

The returned object stores both the public request and the implementation that
actually ran.

```{r nn-metadata}
c(
    requested_backend = attr(nearest, "requested_backend"),
    requested_method = attr(nearest, "requested_method"),
    backend_used = attr(nearest, "backend_used"),
    metric = attr(nearest, "metric"),
    distance_type = attr(nearest, "distance_type")
)
```

## Automatic method selection

The default `method = "auto"` uses compiled shape-, metric-, `k`-, backend-,
and target-recall-aware rules. These rules are deterministic; they do not run a
slow tuning pass inside ordinary user calls.

```{r nn-auto}
auto_nearest <- nn(
    x,
    k = 8,
    backend = "cpu",
    method = "auto",
    metric = "euclidean",
    exclude_self = TRUE,
    n_threads = 2
)

c(
    backend_used = attr(auto_nearest, "backend_used"),
    requested_method = attr(auto_nearest, "requested_method")
)
```

For large two- or three-dimensional self-KNN, `method = "auto"` can select the
native grid route. CPU, CUDA, and Metal have separate implementations. The
explicit CPU call below demonstrates the API on a small two-dimensional matrix.

```{r grid}
x2 <- matrix(runif(2000), ncol = 2)
grid_nearest <- nn(
    x2,
    k = 10,
    backend = "cpu",
    method = "grid",
    metric = "euclidean",
    exclude_self = TRUE,
    n_threads = 2
)

c(
    backend_used = attr(grid_nearest, "backend_used"),
    k = attr(grid_nearest, "k")
)
```

On Apple hardware, the same exact spatial search can run in native Metal. It
keeps the float32 coordinate, grid, candidate, and top-k buffers on the GPU and
returns only the result matrices. CUDA uses its native grid implementation.
Explicit device calls fail when unavailable and never report a CPU fallback as
GPU work.

```{r metal-grid, eval=FALSE}
metal_nearest <- nn(
    x2,
    k = 10,
    exclude_self = TRUE,
    backend = "metal",
    method = "grid",
    metric = "euclidean"
)

cuda_nearest <- nn(
    x2,
    k = 10,
    exclude_self = TRUE,
    backend = "cuda",
    method = "grid",
    metric = "euclidean"
)
```

# Metrics

The public metric set is intentionally small:

- `euclidean`;
- `cosine`;
- `correlation`;
- `inner_product`.

Cosine normalizes each row before searching. Correlation first centers each row
and then applies row normalization. Inner product ranks neighbours by larger
raw dot product, while returned distances keep faissR's smaller-is-better
convention.

```{r metrics}
metric_results <- lapply(
    c("euclidean", "cosine", "correlation", "inner_product"),
    function(metric) {
        out <- nn(
            x,
            k = 5,
            backend = "cpu",
            method = "exact",
            metric = metric,
            exclude_self = TRUE,
            n_threads = 2
        )
        data.frame(
            metric = metric,
            backend_used = attr(out, "backend_used"),
            first_distance = round(out$distances[1, 1], 4)
        )
    }
)

do.call(rbind, metric_results)
```

# Float32 input and output

FAISS and cuVS consume float32 data internally. Ordinary R numeric matrices are
converted once for those compiled routes. If the optional `float` package is
installed, `faissR` can also accept `float::fl()` input and return float32
distance matrices with `output = "float"`.

```{r float32, eval=requireNamespace("float", quietly = TRUE)}
xf <- float::fl(x)

float_nearest <- nn(
    xf,
    k = 5,
    backend = "cpu",
    method = "exact",
    metric = "euclidean",
    exclude_self = TRUE,
    output = "float",
    n_threads = 2
)

c(
    input_type = attr(float_nearest, "input_type"),
    distance_type = attr(float_nearest, "distance_type")
)
class(float_nearest$distances)
```

# Reusing fitted indexes with kNN models

`knn()` can fit a model for repeated prediction, or it can fit and predict in
one call when `Xtest` is supplied. Fitted FAISS Flat, HNSW, IVF, and IVFPQ
indexes are reused by `predict()` where supported.

```{r knn-fit}
fit <- knn(
    Xtrain = x,
    Ytrain = groups,
    k = 7,
    backend = "cpu",
    method = "exact",
    metric = "euclidean",
    n_threads = 2
)

pred <- predict(fit, x[1:6, , drop = FALSE])
prob <- predict(fit, x[1:6, , drop = FALSE], type = "prob")

pred
round(prob, 3)
```

Immediate prediction uses the same interface:

```{r knn-immediate}
immediate <- knn(
    Xtrain = x,
    Ytrain = groups,
    Xtest = x[1:6, , drop = FALSE],
    k = 7,
    backend = "cpu",
    method = "exact",
    metric = "euclidean",
    n_threads = 2
)

immediate
```

# k-means

`fast_kmeans()` uses the same backend style as the nearest-neighbour functions.
The CPU route is available everywhere; CUDA is optional.

```{r kmeans}
km <- fast_kmeans(
    x,
    centers = 3,
    backend = "cpu",
    n_init = 2,
    max_iter = 20
)

c(
    backend = km$backend,
    backend_library = km$backend_library,
    converged = km$converged
)
table(km$cluster, groups)
```

# Optional CUDA use

CUDA-specific calls should be guarded with `cuda_available()` or inspected with
`backend_info()`. The example below is not evaluated on CPU-only builders.

```{r cuda-example, eval=FALSE}
if (cuda_available()) {
    gpu_knn <- nn_gpu(
        x,
        k = 15,
        exclude_self = TRUE,
        method = "auto",
        metric = "euclidean",
        tuning = "auto",
        target_recall = 0.99
    )

    gpu_knn
    host_copy <- gpu_knn_to_host(gpu_knn)
}
```

`nn_gpu()` is for downstream CUDA consumers. It keeps result buffers resident on
the GPU and copies them back to R only when `gpu_knn_to_host()` is called
explicitly.

# Session information

```{r session-info}
sessionInfo()
```
