---
title: "faissR: nearest-neighbor search for Bioconductor workflows"
author: "Stefano Cacciatore"
package: faissR
output:
    BiocStyle::html_document:
        toc: true
        toc_float: true
vignette: >
    %\VignetteIndexEntry{faissR nearest-neighbor workflow}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 6,
    fig.height = 4
)
```

<style>
@media print {
    #TOC,
    .tocify,
    .main-container > .row > .col-sm-4 {
        display: none !important;
    }
    .toc-content {
        margin-left: 0 !important;
        width: 100% !important;
    }
    body {
        font-size: 10pt;
    }
    pre {
        font-size: 7.5pt;
        white-space: pre-wrap;
        overflow-wrap: anywhere;
    }
    table {
        font-size: 8pt;
    }
}
</style>

# Introduction

`faissR` provides exact and approximate nearest-neighbor search for dense
row-wise matrices. A functional installation uses the FAISS C++ library on
CPU; optional CUDA, FAISS GPU, and NVIDIA cuVS providers support GPU execution.
The same metric and result conventions are used through the public R API.

This vignette follows one practical workflow:

1. load a dense single-cell representation;
2. inspect the installed capabilities;
3. run exact and HNSW searches;
4. compare observed HNSW recall with an exact reference;
5. reuse an index for prediction; and
6. preflight input and metric edge cases.

All computational chunks are evaluated. CUDA code is evaluated conditionally
and prints a skip message when the package was built without CUDA support.
Detailed build instructions are available in
`vignette("installation", package = "faissR")`.

# Example data

The installed example is derived from the public Zeisel mouse-brain
single-cell RNA-sequencing data distributed by `scRNAseq`. It contains 50 PCA
coordinates and cell annotations for 3,005 cells, but no expression matrix.
The source URLs, complete preprocessing procedure, seed, and package versions
are recorded by `inst/scripts/prepare_zeisel_brain_pca.R` and documented in
`?zeisel_brain_pca`.

```{r data, message=FALSE, warning=FALSE}
library("faissR")
library("SingleCellExperiment")

example_file <- system.file(
    "extdata", "zeisel_brain_pca.rds", package = "faissR"
)
sce <- readRDS(example_file)
x <- reducedDim(sce, "PCA")
storage.mode(x) <- "double"
groups <- droplevels(factor(colData(sce)$level1class))

dim(x)
head(table(groups), 6L)
```

Nearest-neighbor search should normally receive a dense low-dimensional
representation such as PCA scores, not a full sparse gene-by-cell count
matrix. The following table estimates the principal host allocations for 50
PCs and `k = 30` without allocating the larger matrices.

```{r memory-budget}
cells <- c(3005, 100000, 1000000)
data.frame(
    cells = cells,
    input_double_MiB = round(8 * cells * 50 / 1024^2, 1),
    native_float32_MiB = round(4 * cells * 50 / 1024^2, 1),
    result_MiB = round((4 + 8) * cells * 30 / 1024^2, 1)
)
```

The result estimate includes int32 identifiers and R double distances. Exact
full self-search also has quadratic computational cost, so large studies will
usually require an approximate method, an exact validation subsample, and an
explicit memory budget.

# Inspect execution capabilities

Hardware visibility and package capability are different questions. The
terminology here follows the framework-independent distinction implemented by
the CRAN package [`gpuinfo`](https://cran.r-project.org/package=gpuinfo): a
visible GPU or driver does not prove that a consuming R package was compiled
with a usable GPU provider.

In `faissR`:

- `faissR_backend()` reads or sets the requested session device;
- `backend_info()` reports providers compiled into this installation and
    whether their runtimes are currently usable;
- `nn_capabilities(runtime = TRUE)` checks a method, device, and metric
    combination; and
- `cuda_available()` is package-specific: it requires both a CUDA-enabled
    `faissR` build and a runtime-visible device.

`faissR` uses `gpuinfo` for platform-level hardware diagnosis. Package-specific
provider checks remain authoritative for deciding whether a `faissR` call can
execute.

```{r backend-status}
faissR_backend()

hardware <- gpuinfo::hardware_info()
hardware$gpu

backend_info()[, c(
    "backend", "available", "device"
)]

cpu_euclidean <- subset(
    nn_capabilities(runtime = TRUE),
    backend == "cpu" & metric == "euclidean"
)
cpu_euclidean[, c(
    "method", "supported", "runtime_available", "resolved_backend"
)]
```

CPU is the default. An explicit function argument has highest precedence,
followed by `options(faissR.backend)`, `FAISSR_BACKEND`, and finally CPU.

```{r backend-precedence}
old_backend <- faissR_backend("cpu")
stopifnot(identical(faissR_backend(), "cpu"))
```

An explicit `backend = "cuda"` never silently falls back to CPU. `faissR`
does not provide a Metal backend.

# Run a nearest-neighbor search

Use `exclude_self = TRUE` when the query and reference are the same rows and a
row must not return its own identifier.

```{r exact-search}
exact_time <- system.time(
    exact <- nn(
        x,
        k = 15,
        backend = "cpu",
        method = "exact",
        metric = "euclidean",
        exclude_self = TRUE,
        n_threads = 2
    )
)[["elapsed"]]

exact
dim(exact$indices)
head(exact$indices, 3L)
head(round(exact$distances, 3), 3L)
```

Indices are one-based R integers. Distances are sorted within each query row.
The returned metadata distinguishes the user request from the provider that
actually executed it.

```{r result-metadata}
c(
    requested_backend = attr(exact, "requested_backend"),
    requested_method = attr(exact, "requested_method"),
    backend_used = attr(exact, "backend_used"),
    metric = attr(exact, "metric"),
    exact = attr(exact, "exact"),
    exclude_self = attr(exact, "exclude_self"),
    input_type = attr(exact, "input_type")
)
```

`exact`, `flat`, and `bruteforce` request exhaustive mathematics but express
different provider intentions:

| Method | Practical use |
|---|---|
| `exact` | Let `faissR` choose an available exhaustive provider. |
| `flat` | Require a FAISS Flat provider. |
| `bruteforce` | Prefer direct cuVS on CUDA; use FAISS Flat on CPU. |

The `backend_used` attribute is the authoritative resolved implementation.

# Approximate search and observed recall

HNSW trades exactness for speed and index storage. `tuning = "auto"` chooses
parameters for the resolved method. `target_recall` accepts the calibrated
tiers 0.90, 0.95, and 0.99; it is not interpolated. For a new application,
measure observed recall on representative queries against an exact reference.

```{r hnsw-search}
hnsw_time <- system.time(
    hnsw <- nn(
        x,
        k = 15,
        backend = "cpu",
        method = "hnsw",
        metric = "euclidean",
        tuning = "auto",
        target_recall = 0.99,
        exclude_self = TRUE,
        n_threads = 2
    )
)[["elapsed"]]

query_recall <- vapply(
    seq_len(nrow(exact$indices)),
    function(i) mean(hnsw$indices[i, ] %in% exact$indices[i, ]),
    numeric(1L)
)

c(
    elapsed_seconds = unname(hnsw_time),
    mean_recall = mean(query_recall),
    fifth_percentile_recall = unname(quantile(query_recall, 0.05)),
    minimum_recall = min(query_recall),
    backend_used = attr(hnsw, "backend_used")
)
```

# Automatic selection

The execution device is selected with `backend = "cpu"` or
`backend = "cuda"`. If `backend` is omitted, faissR follows the package option,
then `FAISSR_BACKEND`, and finally uses CPU. Two other arguments control
different decisions:

| Argument | Decision |
|---|---|
| `method = "auto"` | choose a method family |
| `tuning = "auto"` | choose parameters within that method |

`method = "auto"` is a static, calibration-informed experimental policy. It
does not benchmark the current data during an ordinary call. CPU automatic
selection has less independent validation than the CUDA policy; use an
explicit method when reproducibility, memory, or latency constraints require a
fixed route.

```{r auto-selection}
auto_result <- nn(
    x,
    k = 8,
    backend = "cpu",
    method = "auto",
    metric = "euclidean",
    exclude_self = TRUE,
    n_threads = 2
)

c(
    requested_method = attr(auto_result, "requested_method"),
    selected_method = attr(auto_result, "auto_selection")$predicted_method,
    backend_used = attr(auto_result, "backend_used"),
    policy_status = attr(auto_result, "auto_policy_status")
)
```

# Metrics and edge cases

The public metric set is `"euclidean"`, `"cosine"`, and `"correlation"`.
Euclidean results are ordinary L2 distances, not squared values. Cosine
normalizes rows; correlation centers and then normalizes rows. All routes
reject `NA`, `NaN`, and infinite input.

Zero vectors have undefined cosine distance, and constant rows have undefined
correlation distance. CPU routes apply documented finite conventions; CUDA
routes reject those rows because no CPU repair is performed. Use
`nn_metric_preflight()` before an expensive call to identify affected
one-based row indices and the backend action.

```{r metric-preflight}
edge_x <- rbind(c(0, 0), c(1, 0), c(1, 1))
nn_metric_preflight(
    edge_x,
    metric = "cosine",
    backend = "cuda"
)
```

Unsupported method, backend, and metric combinations fail explicitly. The
capability table can be used to filter a workflow before execution.

```{r unsupported-combination}
tryCatch(
    nn(x, k = 5, backend = "cpu", method = "cagra"),
    error = function(e) conditionMessage(e)
)
```

# Dense input contract

`faissR` searches dense, two-dimensional numeric data only. Accepted inputs
are numeric matrices, numeric data frames, and optional `float::fl()`
matrices, with observations in rows. It does not implement sparse nearest-
neighbor search. Sparse, delayed, file-backed, and other matrix-like objects
are rejected before conversion and are never silently materialized with
`as.matrix()`.

```{r sparse-input-guard}
sparse_example <- Matrix::Matrix(diag(4), sparse = TRUE)
tryCatch(
    nn(sparse_example, k = 1),
    error = function(e) conditionMessage(e)
)
```

For single-cell data, keep the expression or count assay sparse during
normalization and dimension reduction, then pass a dense, size-checked cell
embedding such as PCA coordinates to `faissR`. If direct materialization is
appropriate for another workflow, it must be performed explicitly by the
caller before the `faissR` call.

Ordinary R double matrices are converted internally to contiguous float32
storage for FAISS and cuVS. This numeric conversion applies only after the
input has passed the dense-input check. The original R object remains live
during conversion, so the double source, float32 payload, index workspace,
and result must all fit in host memory.

# Reuse an index for prediction

`knn()` creates a fitted classification or regression model. Supported FAISS
indexes are retained for later `predict()` calls in the same R session. Native
indexes are session-local external pointers: after `saveRDS()` and
`readRDS()`, the R model remains usable but the native index is rebuilt on the
first prediction.

```{r fitted-model}
build_time <- system.time(
    fit <- knn(
        Xtrain = x,
        Ytrain = groups,
        k = 7,
        backend = "cpu",
        method = "exact",
        metric = "euclidean",
        n_threads = 2
    )
)[["elapsed"]]

query_time <- system.time(
    prediction <- predict(fit, x[1:12, , drop = FALSE])
)[["elapsed"]]

c(
    build_seconds = unname(build_time),
    query_seconds = unname(query_time),
    query_source = attr(prediction, "faissR_nn")$query_source,
    backend_used = attr(prediction, "faissR_nn")$resolved_backend
)
head(as.vector(prediction))
```

The imported `Biobase` package is supported for conventional
`ExpressionSet` workflows. Samples are stored in columns, so transpose the
assay before search.

```{r biobase-example}
data("sample.ExpressionSet", package = "Biobase")
biobase_x <- t(Biobase::exprs(sample.ExpressionSet)[seq_len(32L), ])
biobase_result <- nn(
    biobase_x,
    k = 3,
    backend = "cpu",
    method = "exact",
    exclude_self = TRUE
)
dim(biobase_result$indices)
```

# Optional CUDA execution

Guard CUDA work with the package-specific `cuda_available()` predicate. The
conditional code below is evaluated on every builder; only the GPU operation
is skipped when CUDA is unavailable.

```{r cuda-example}
if (cuda_available()) {
    cuda_result <- nn(
        x,
        k = 15,
        backend = "cuda",
        method = "exact",
        metric = "euclidean",
        exclude_self = TRUE
    )
    c(
        backend_used = attr(cuda_result, "backend_used"),
        device = attr(cuda_result, "device")
    )
} else {
    cat(
        "CUDA example skipped: this faissR installation has ",
        "no usable CUDA route."
    )
}
```

`nn_gpu()` serves a different workflow: it returns an owning
`faissR_gpu_knn` object whose neighbor buffers stay on the device until
`gpu_knn_to_host()` is called. This is intended for compiled downstream CUDA
consumers. The object must remain alive while non-owning device pointers are in
use and cannot be serialized across sessions or devices. See `?nn_gpu` and
`?faissR_c_api` for the ownership and C-callable interface contracts.

`candidate_knn()` and `fast_kmeans()` are secondary interfaces that reuse the
same metric and backend infrastructure. Their complete contracts and examples
are available in `?candidate_knn` and `?fast_kmeans`.

# Session information

```{r session-info}
sessionInfo()
```
